Galaxy image statistical generator: how to be confident?

### 2024-09-11 paper/main.tex
experiemnt

\title{Galaxy image statistical generator: how to be confident?}
\author{Jean-Eric Campagne}
\address{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\email{jean-eric.campagne@ijclab.in2p3.fr}
\date{\today}
\calclayout  % switch off left right margin different even/odd pages.

@@

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g. deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.

@@

We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi}.

Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray level is coded using 5-bits precision.
{\color{red}The training dataset of $250,000$ galaxies is available in github}. With such dataset,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size the range $10^4-10^5$.

The training

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{aa}%{elsarticle-harv}%%{mnras}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Details on the dataset}

>>>>

\title{Galaxy image statistical generator: how to be confident?}
\author{Jean-Eric Campagne}
\address{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\email{jean-eric.campagne AT ijclab.in2p3.fr}
\date{\today}
\calclayout  % switch off left right margin different even/odd pages.

@@

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks as deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and what we are concern in this article the synthetic galaxy image generations  going beyond parametrized analytic light profile simulations (e.g. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.

@@

We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size in the range $10^4-10^5$.

So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{aa}%{apalike}%{elsarticle-harv}%%{mnras}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Details on the dataset}
